---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "YOUR NAME HERE"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  # Keep the unfinished template display-only until its placeholders are filled.
  eval: false
  warning: false
  message: false
---

::: {.callout-important}
## Read this first, then delete this box

**Where this file goes:** `assignments/assignment2.qmd` in your portfolio repo.

**This is a portfolio piece.** It will be published on your website where employers
can see it. So:

- **Delete every callout box like this one** before your final render. They're
  instructions, not content.
- Replace `YOUR NAME HERE` in the header above.
- Write in complete sentences. Someone who isn't in this class should be able to
  read it and follow what you did.

**Graded 0 / 1 / 2.** A 2 means it runs, it's complete, and the writing shows you
understood what the numbers meant — not that you did something fancy.

**Due:** before next class. Rendered, committed, pushed, and visible on your live site.
:::

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

::: {.callout-note}
## Delete this box

If this chunk fails, your Census API key isn't set up. Check your `.Renviron`
file, then **restart R** — that's the step people forget.
:::

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

::: {.callout-tip}
## Delete this box

Rewrite the paragraph below in your own words once you've picked your state and
variable. Two or three sentences saying what you're looking at and why.
:::

**In this analysis, I look at [YOUR VARIABLE] across counties in [YOUR STATE], and
ask which communities we measure most and least reliably.**

---

## Choosing what to measure

::: {.callout-note}
## Instructions — delete this box

Pick **a state** and **a variable**. Anything except Pennsylvania and median
household income, which we did in class.

Some variables that work well:

| Code | What it is |
|------|-----------|
| `B19013_001` | Median household income |
| `B25064_001` | Median gross rent |
| `B25077_001` | Median home value |
| `B01002_001` | Median age |
| `B08303_001` | Total commuters (travel time table) |
| `B15003_022` | Number with a bachelor's degree |

Want something else? Run this in your **console** (not in a chunk):

```r
vars <- load_variables(2023, "acs5")
View(vars)
```

Then use the search box.
:::

```{r}
my_state <- "___"        # two-letter abbreviation, in quotes
my_variable <- "___"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

::: {.callout-tip}
## Delete this box

Below, say how many counties are in your state and confirm your row count matches.
One sentence.
:::

**[How many rows did you get? Does that match the number of counties in your state?]**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = ___ / ___ * 100)
```

::: {.callout-note}
## Delete this box

Fill in the two blanks. You want: how big is the margin of error, relative to the
estimate, as a percentage?
:::

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

::: {.callout-tip}
## Delete this box

Write a sentence or two naming the least reliable county and its `moe_pct`.
:::

**[Which county has the largest margin of error relative to its estimate? How large
is it?]**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

::: {.callout-note}
## Delete this box

Those cutoffs (5% and 10%) are a bit of a judgment call, not a law of nature. If almost
every county lands in one category, adjust them so the split is informative for
*your* variable — then say in your writing that you changed them and why. (I have a feeling for county-level, you may end up in high confidence everywhere - is smaller geographies where it really shows up. So feel free to change those numbers around a bit)
:::

**[How many counties fall into each category?]**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

::: {.callout-note}
## Delete this box

Notice the column names: `myvarE`, `myvarM`, `popE`, `popM`. The `E` and `M`
suffixes are estimate and margin of error.
:::

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(___) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(___)
  )
```

::: {.callout-note}
## Delete this box

Two blanks. What are you grouping by? What are you averaging?
:::

::: {.callout-tip}
## Delete this box

This table is the heart of the assignment. Describe what it shows in two or three
sentences.
:::

**[What is the average population of high-confidence counties versus low-confidence
counties? Is there a pattern?]**

---

## What this means for policy

::: {.callout-important}
## Instructions — delete this box

**Write 200–250 words.** 

Address all three:

1. **Which communities in your state do we know the least about?** Name specific
   counties from your results.
2. **Why?** Connect the pattern to how the ACS actually works — it's a sample, and
   smaller places yield smaller samples.
3. **What could go wrong?** Imagine a state agency allocating funding using your
   variable, treating every county's number as equally certain. Who is likely to be
   hurt, and how?

Use your own numbers. 
:::

**[Your 200–250 word reflection goes here.]**

---

## Optional challenges - For those craving more

::: {.callout-note}
## Delete this box

**None of these are graded.** There are no extra points. A complete, well-written assignment earns a 2 either way.

They're here for anyone who finished the above and found it straightforward. Pick
one (or more) that interests you. 

1. **Propagate the margin of error.** If you compute a derived value — a percentage,
   a ratio, a sum — its uncertainty depends on the uncertainty of its inputs, and
   you can't just carry the original MOE along. The Census Bureau publishes formulas
   for this, and `tidycensus` implements them: `moe_sum()`, `moe_prop()`,
   `moe_ratio()`. Compute something derived, and get its margin of error right.

2. **Parameterize the report.** Add `params:` to the YAML so the state is an input
   rather than hard-coded, and the same file regenerates for any state. Render it
   from the terminal with `quarto render assignment2.qmd -P state:NY`.

3. **Write a function and iterate.** Wrap the data pull in a function, then use
   `purrr::map_dfr()` to run it across several states at once. Does the
   small-places-are-noisier pattern hold everywhere, or are some states different?

4. **Change the geographic scale.** Run the same variable at `geography = "tract"`
   inside one county. Compare the margins of error to what you saw at county level.
   What does that imply about the resolution at which you can actually trust ACS
   data?

If you try one, add a short section describing what you did and what you found.
:::

---

## Before you submit

::: {.callout-note}
## Delete this box — checklist

- [ ] Your name is in the YAML header
- [ ] Every instruction callout is deleted
- [ ] Every `[bracketed prompt]` is replaced with your own writing
- [ ] No `___` blanks remain anywhere
- [ ] The file is at `assignments/assignment2.qmd`
- [ ] **Session → Restart R**, then render the whole project. It works.
- [ ] Committed and pushed
- [ ] You opened your live site and looked at the published page
:::
