---
title: "Assignment3: Verify the Claim: Poverty Change in Philadelphia"
author: "Chenyan Zhao"
date: September 17, 2026
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
---

## Part 1 Reconstruct

### Article and analytical scope

I analyze Alfred Lubrano's 2018 WHYY / Philadelphia Inquirer article, “New
Census figures on Philly neighborhoods show inequality, high numbers of whites
living in poverty.” **The article reports that Eastwick's poverty rate increased
from 12 percent in the 2006–2010 ACS period to 26 percent in the 2013–2017 ACS
period. It also states that the newer figures come from the American Community
Survey (ACS) five-year estimates.**

The article reports results for Philadelphia neighborhoods, but neighborhoods
such as Eastwick are not official Census geographies. The Census Bureau
publishes ACS data for Census tracts, which have GEOIDs and defined boundaries.
Since the article does not provide its tract-to-neighborhood crosswalk or
boundary file, I cannot exactly reproduce its neighborhood-level estimates.
Therefore, I reconstruct and test the article's general poverty-change claim
at the Census-tract level for Philadelphia County, Pennsylvania.

Source article:
<https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/>

### Variables

I use ACS detailed table `B17001`, *Poverty Status in the Past 12 Months by
Sex by Age*. The table provides a poverty count and its margin of error, as
well as the population for whom poverty status is determined.

- `B17001_001`: Total population for whom poverty status is determined. I use
  this as the denominator.
- `B17001_002`: Population with income in the past 12 months below the poverty
  level. I use this as the numerator.

For each Census tract, the poverty rate will be calculated as:

$$
\text{Poverty rate} =
\frac{\text{B17001\_002}}{\text{B17001\_001}}
$$

These variables match the article's general discussion of overall poverty
rates. They do not measure poverty only for a specific age group, sex, or race.
The article also discusses white poverty in Holmesburg, but that would require
a separate race-specific analysis and a clear definition of “white” that the
article does not provide. I therefore focus this analysis on overall poverty.

```{r}
#| include: false

library(tidycensus)
library(tidyverse)
library(dplyr)
library(gt)
```

```{r}
#| label: pull-acs-data
#| message: false
#| warning: false

# Get data
get_poverty <- function(yr) {
  get_acs(
    geography = "tract",
    variables = c(
      below_poverty = "B17001_002",
      total = "B17001_001"
    ),
    state = "PA",
    county = "Philadelphia",
    year = yr,
    survey = "acs5",
    output = "wide"
  ) %>%
    mutate(period = paste0(yr - 4, "-", yr)
  )
}

# Retrieve and combine tract-level poverty data.
poverty_change <- bind_rows(
  get_poverty(2010),
  get_poverty(2017)
)
```

```{r}
#| label: data-coverage
#| tbl-cap: "Table 1. Census tracts retrieved from ACS 5-year data"

poverty_change %>%
  count(period, name = "Number of Census tracts") %>%
  knitr::kable(
    col.names = c("ACS period", "Number of Census tracts")
  )
```

### Years, survey, and assumptions

I use the 2010 and 2017 ACS five-year releases because they match the two periods discussed in the article: 2006–2010 and 2013–2017. I use `acs5` because tract-level data are available in the five-year ACS, while the one-year ACS does not provide tract estimates.

The article does not explain exactly which Census table, poverty-rate denominator, or neighborhood boundaries it used. I therefore use table `B17001` and calculate poverty rates for Census tracts in Philadelphia County. This cannot exactly reproduce the article’s neighborhood results because neighborhoods do not have official Census boundaries. I also match tracts by `GEOID` before comparing the two periods.

The two ACS periods do not overlap, so they do not use the same survey years. However, each estimate still represents a five-year average. For this reason, the results show changes in estimated tract poverty rates, not changes experienced by the same individual residents.

## Part 2 — Test

### Check missing values

First, I calculate poverty rates and their margins of error for all returned
tract-period observations. 

```{r}
#| label: calculate-poverty-rates

poverty_rates <- poverty_change %>%
  mutate(
    poverty_rate = below_povertyE / totalE,
    poverty_rate_moe = moe_prop(
      below_povertyE,
      totalE,
      below_povertyM,
      totalM
    )
  )
```

I then check which Census tracts have missing poverty-rate results in either
ACS period.

```{r}
#| label: inspect-missing-values

missing_tracts <- poverty_rates %>%
  filter(
    is.na(poverty_rate) |
    is.na(poverty_rate_moe)
  ) %>%
  select(
    GEOID,
    NAME,
    period,
    totalE,
    totalM,
    below_povertyE,
    below_povertyM,
    poverty_rate,
    poverty_rate_moe
  )

missing_tracts
```

The missing results come from 15 tract-period observations: seven in
2006–2010 and eight in 2013–2017. In every case, the estimated population for
whom poverty status was determined (`totalE`) is zero, so a poverty rate cannot
be calculated. Seven tracts have this problem in both periods, while Census
Tract 9806 has a usable estimate in 2006–2010 but not in 2013–2017. I do not
treat these missing rates as zero. I exclude them because zero means there is
no estimated population in the poverty universe, not that the poverty rate is
zero.

```{r}
#| label: remove-invalid-rates

poverty_rates_clean <- poverty_rates %>%
  filter(
    !is.na(poverty_rate),
    !is.na(poverty_rate_moe)
  )
```

### Reliability

I calculate the coefficient of variation using the poverty-rate margin of
error. Following Jurjevich et al. (2018), estimates with a CV below 12 percent
are classified as reliable; estimates from 12 to 40 percent are somewhat
reliable; and estimates above 40 percent are unreliable.

```{r}
#| label: calculate-reliability

poverty_rates_clean <- poverty_rates_clean %>%
  mutate(
    cv = (poverty_rate_moe / 1.645) / poverty_rate * 100,
    reliability = case_when(
      cv < 12 ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE ~ "Unreliable"
    )
  )
```

```{r}
#| label: reliability-summary

poverty_rates_clean %>%
  count(period, reliability)
```

The table reports the reliability of every usable tract estimate in the
2013–2017 ACS period. The color-coded flags encourage cautious interpretation
rather than automatically excluding valid ACS observations.

```{r}
#| label: reliability-table

poverty_rates_clean %>%
  filter(period == "2013-2017") %>%
  mutate(
    reliability_symbol = case_when(
      cv < 12  ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE     ~ "🔴"
    ),
    tract = str_remove(NAME, ", Philadelphia County, Pennsylvania"),
    rate = poverty_rate,
    rate_moe = poverty_rate_moe
  ) %>%
  arrange(desc(cv)) %>%
  select(reliability_symbol, tract, rate, rate_moe, cv, reliability) %>%
  gt() %>%
  tab_header(
    title = "Reliability of Philadelphia tract poverty-rate estimates",
    subtitle = "2013-2017 ACS five-year estimates"
  ) %>%
  cols_label(
    reliability_symbol = "",
    tract = "Census tract",
    rate = "Poverty rate",
    rate_moe = "MOE",
    cv = "CV (%)",
    reliability = "Reliability"
  ) %>%
  fmt_percent(columns = rate, decimals = 1) %>%
  fmt_percent(columns = rate_moe, decimals = 1) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_align(align = "center", columns = reliability_symbol) %>%
  tab_source_note(
    "Source: U.S. Census Bureau ACS five-year estimates via tidycensus. Jurjevich et al. (2018) thresholds: CV < 12% reliable; 12-40% somewhat reliable; > 40% unreliable."
  )%>%
  tab_options(
    # keep the sheet into a box, so that it will be easier to read.
    container.height = px(500),
    container.overflow.y = TRUE,
    container.width = pct(100),
    container.overflow.x = TRUE
  )
```

### Change over time

Next, I reshape the data so that each tract has one row containing poverty
rates and margins of error for both ACS periods.

```{r}
#| label: reshape-poverty-data

poverty_wide <- poverty_rates_clean %>%
  select(
    GEOID,
    NAME,
    period,
    poverty_rate,
    poverty_rate_moe
  ) %>%
  pivot_wider(
    names_from = period,
    values_from = c(poverty_rate, poverty_rate_moe)
  )
```

I identify tracts that have usable data in only one period. Those tracts cannot be included in the change test because a difference requires valid estimates and margins of error for both periods.

```{r}
#| label: inspect-incomplete-comparisons

incomplete_tracts <- poverty_wide %>%
  filter(
    is.na(`poverty_rate_2006-2010`) |
    is.na(`poverty_rate_2013-2017`) |
    is.na(`poverty_rate_moe_2006-2010`) |
    is.na(`poverty_rate_moe_2013-2017`)
  )

incomplete_tracts
```

The ACS reports 90 percent margins of error. I divide each MOE by 1.645 to calculate its standard error, combine the two standard errors to calculate the standard error of the difference, and multiply by 1.645 to obtain the 90 percent margin of error for the change.

```{r}
#| label: test-poverty-rate-change

poverty_tested <- poverty_wide %>%
  filter(
    !is.na(`poverty_rate_2006-2010`),
    !is.na(`poverty_rate_2013-2017`),
    !is.na(`poverty_rate_moe_2006-2010`),
    !is.na(`poverty_rate_moe_2013-2017`)
  ) %>%
  mutate(
    change = `poverty_rate_2013-2017` -
      `poverty_rate_2006-2010`,

    se_1 = `poverty_rate_moe_2006-2010` / 1.645,
    se_2 = `poverty_rate_moe_2013-2017` / 1.645,

    se_diff = sqrt(se_1^2 + se_2^2),
    moe_diff = se_diff * 1.645,

    z = change / se_diff,
    significant = abs(z) > 1.645,

    change_direction = case_when(
      significant & change > 0 ~ "Significant increase",
      significant & change < 0 ~ "Significant decrease",
      TRUE ~ "Not distinguishable from zero"
    )
  )

change_counts <- poverty_tested %>%
  count(change_direction)

change_counts
```

Of the `r nrow(poverty_tested)` Census tracts with usable estimates in both
periods, `r sum(poverty_tested$significant)` show a statistically detectable
poverty-rate change at the 90 percent confidence level. The remaining
`r sum(!poverty_tested$significant)` tracts do not show a change that is
statistically distinguishable from zero.

```{r}
#| label: change-summary-plot
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Most Philadelphia census tracts did not have a statistically detectable change in poverty rate."

ggplot(change_counts, aes(x = change_direction, y = n, fill = change_direction)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.15) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Significant increase" = "#C85A5A",
      "Significant decrease" = "#4F81BD",
      "Not distinguishable from zero" = "#8A8A8A"
    )
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Results of the tract-level poverty-rate change test",
    subtitle = "90% confidence tests for 376 matched Philadelphia census tracts",
    x = NULL,
    y = "Number of census tracts",
    caption = "Source: U.S. Census Bureau, ACS five-year estimates via tidycensus."
  ) +
  theme_minimal()
```

```{r}
#| label: change-error-bar-plot
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Largest statistically detectable tract-level poverty-rate changes, with 90% margins of error."

change_plot_data <- poverty_tested %>%
  filter(significant) %>%
  arrange(desc(abs(change))) %>%
  slice_head(n = 20) %>%
  mutate(
    tract = str_remove(NAME, ", Philadelphia County, Pennsylvania"),
    change_pp = change * 100,
    moe_diff_pp = moe_diff * 100,
    direction = case_when(
      change > 0 ~ "Increase",
      TRUE ~ "Decrease"
    )
  )

ggplot(change_plot_data, aes(x = reorder(tract, change_pp), y = change_pp)) +
  geom_col(aes(fill = direction), width = 0.7) +
  geom_errorbar(
    aes(ymin = change_pp - moe_diff_pp, ymax = change_pp + moe_diff_pp),
    width = 0.25
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  scale_fill_manual(values = c("Increase" = "#C85A5A", "Decrease" = "#4F81BD")) +
  labs(
    title = "Largest statistically detectable poverty-rate changes",
    subtitle = "Twenty largest absolute changes, 2006-2010 versus 2013-2017",
    x = NULL,
    y = "Change in poverty rate (percentage points)",
    fill = NULL,
    caption = "Error bars show 90% margins of error; dashed line marks no change. Source: U.S. Census Bureau, ACS five-year estimates via tidycensus."
  ) +
  theme_minimal()
```

## Part 3 — Adjudicate

**Supported.** The tract results support the article's broader point that
poverty worsened in some parts of Philadelphia. Of the 376 tracts that could
be compared, 36 had a statistically detectable increase in poverty.
**Not supported.** The results do not support the idea that poverty increased
across the city as a whole: 27 tracts had a detectable decrease, and 313 did
not show a clear change. The pattern was mixed rather than citywide.

**Cannot be determined.** I cannot verify the article's exact figures for
Eastwick (12% to 26%), Fairhill (61%), or white poverty in Holmesburg. The
article does not publish the neighborhood boundaries or the tracts used in
those calculations. The Holmesburg claim would also require a race-specific
poverty measure, while this analysis measures overall poverty.

## Part 4 — Rewrite

New American Community Survey data show that poverty changed unevenly across
Philadelphia between 2006–10 and 2013–17. Among 376 census tracts with
comparable estimates, 36 had a clear increase in poverty and 27 had a clear
decrease. In the remaining 313 tracts, the estimated changes were too small
relative to survey uncertainty to draw a firm conclusion. The results show
worsening poverty in some parts of the city and declines in others, while the
direction of change remains unclear across most tracts. The article's exact
figures for Eastwick, Fairhill, and Holmesburg cannot be confirmed because the
neighborhood boundaries used in those calculations were not published.
