---
title: "Assignment3: Verify the Claim: Poverty Change in Philadelphia"
author: "Chenyan Zhao"
date: September 17, 2026
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
---

## Part 1 Reconstruct

### Article and analytical scope

I analyze Alfred Lubrano's 2018 WHYY / Philadelphia Inquirer article, “New
Census figures on Philly neighborhoods show inequality, high numbers of whites
living in poverty.” The article reports that Eastwick's poverty rate increased
from 12 percent in the 2006–2010 ACS period to 26 percent in the 2013–2017 ACS
period. It also states that the newer figures come from the American Community
Survey (ACS) five-year estimates.

The article reports results for Philadelphia neighborhoods, but neighborhoods
such as Eastwick are not official Census geographies. The Census Bureau
publishes ACS data for Census tracts, which have GEOIDs and defined boundaries.
Since the article does not provide its tract-to-neighborhood crosswalk or
boundary file, I cannot exactly reproduce its neighborhood-level estimates.
Therefore, I reconstruct and test the article's general poverty-change claim
at the Census-tract level for Philadelphia County, Pennsylvania.

Source article:
<https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/>

### Variables

I use ACS detailed table `B17001`, *Poverty Status in the Past 12 Months by
Sex by Age*. The table provides a poverty count and its margin of error, as
well as the population for whom poverty status is determined.

- `B17001_001`: Total population for whom poverty status is determined. I use
  this as the denominator.
- `B17001_002`: Population with income in the past 12 months below the poverty
  level. I use this as the numerator.

For each Census tract, the poverty rate will be calculated as:

$$
\text{Poverty rate} =
\frac{\text{B17001\_002}}{\text{B17001\_001}}
$$

These variables match the article's general discussion of overall poverty
rates. They do not measure poverty only for a specific age group, sex, or race.
The article also discusses white poverty in Holmesburg, but that would require
a separate race-specific analysis and a clear definition of “white” that the
article does not provide. I therefore focus this analysis on overall poverty.

```{r}
#| include: false

library(tidycensus)
library(tidyverse)
library(dplyr)
library(gt)
```

```{r}
#| label: pull-acs-data
#| message: false
#| warning: false

# Get data
get_poverty <- function(yr) {
  get_acs(
    geography = "tract",
    variables = c(
      below_poverty = "B17001_002",
      total = "B17001_001"
    ),
    state = "PA",
    county = "Philadelphia",
    year = yr,
    survey = "acs5",
    output = "wide"
  ) %>%
    mutate(period = paste0(yr - 4, "-", yr)
  )
}

# Retrieve and combine tract-level poverty data.
poverty_change <- bind_rows(
  get_poverty(2010),
  get_poverty(2017)
)
```

```{r}
#| label: data-coverage
#| tbl-cap: "Table 1. Census tracts retrieved from ACS 5-year data"

poverty_change %>%
  count(period, name = "Number of Census tracts") %>%
  knitr::kable(
    col.names = c("ACS period", "Number of Census tracts")
  )
```

### Years, survey, and assumptions

I use the 2010 ACS five-year release to represent the article's 2006–2010
period, and the 2017 ACS five-year release to represent its 2013–2017 period.
I specify `survey = "acs5"` because the article explicitly identifies its
data as ACS five-year estimates. The five-year ACS is also the appropriate
product for this assignment because it publishes estimates for Census tracts.
The ACS one-year product (`acs1`) does not provide tract-level estimates, so
it cannot support this tract-level comparison.

The article does not state its exact Census table, poverty-rate denominator,
tract-to-neighborhood crosswalk, or statistical testing method. I therefore
make several assumptions. First, I assume its overall poverty-rate figures
correspond to the ACS definition in table `B17001`. Second, I assume that
Philadelphia County is the appropriate geography for the City of Philadelphia.
Third, I use Census tracts as the analysis unit because the article's custom
neighborhood boundaries are unavailable. Finally, I assume that matching
tracts by `GEOID` permits comparison between the 2010 and 2017 ACS releases;
I will check for unmatched GEOIDs before calculating change.

The two ACS periods do **not** overlap. The first period contains data from
2006 through 2010, while the second contains data from 2013 through 2017.
Because they have no years in common, this comparison does not reuse the same
annual ACS samples in both estimates. However, these are still pooled
five-year survey estimates rather than observations of the same individuals
over time. Therefore, the results should be interpreted as changes in
estimated tract-level poverty rates, not as evidence that the same residents
became poorer.

## Part 2 — Test

### Check missing values

First, I calculate poverty rates and their margins of error for all returned
tract-period observations. I do not remove missing values at this stage.

```{r}
#| label: calculate-poverty-rates

poverty_rates <- poverty_change %>%
  mutate(
    poverty_rate = below_povertyE / totalE,
    poverty_rate_moe = moe_prop(
      below_povertyE,
      totalE,
      below_povertyM,
      totalM
    )
  )
```

I then check which Census tracts have missing poverty-rate results in either
ACS period.

```{r}
#| label: inspect-missing-values

missing_tracts <- poverty_rates %>%
  filter(
    is.na(poverty_rate) |
    is.na(poverty_rate_moe)
  ) %>%
  select(
    GEOID,
    NAME,
    period,
    totalE,
    totalM,
    below_povertyE,
    below_povertyM,
    poverty_rate,
    poverty_rate_moe
  )

missing_tracts
```

I do not treat missing values as zero. A poverty rate cannot be calculated
when the poverty-status universe is zero or when the ACS does not provide the
required estimate or margin of error. I therefore exclude only these
tract-period observations from the statistical comparison.

```{r}
#| label: remove-invalid-rates

poverty_rates_clean <- poverty_rates %>%
  filter(
    !is.na(poverty_rate),
    !is.na(poverty_rate_moe)
  )
```

### Reliability

I calculate the coefficient of variation using the poverty-rate margin of
error. Following Jurjevich et al. (2018), estimates with a CV below 12 percent
are classified as reliable; estimates from 12 to 40 percent are somewhat
reliable; and estimates above 40 percent are unreliable.

```{r}
#| label: calculate-reliability

poverty_rates_clean <- poverty_rates_clean %>%
  mutate(
    cv = (poverty_rate_moe / 1.645) / poverty_rate * 100,
    reliability = case_when(
      cv < 12 ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE ~ "Unreliable"
    )
  )
```

```{r}
#| label: reliability-summary

poverty_rates_clean %>%
  count(period, reliability)
```

The table below makes the reliability classification visible for the 15 tracts
with the largest coefficients of variation in the most recent ACS period. I
show the least reliable estimates rather than hiding them because high sampling
uncertainty is directly relevant to interpreting apparent change.

```{r}
#| label: reliability-table

poverty_rates_clean %>%
  filter(period == "2013-2017") %>%
  mutate(
    reliability_symbol = case_when(
      cv < 12 ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE ~ "🔴"
    ),
    tract = str_remove(NAME, ", Philadelphia County, Pennsylvania"),
    rate = poverty_rate,
    rate_moe = poverty_rate_moe
  ) %>%
  arrange(desc(cv)) %>%
  slice_head(n = 15) %>%
  select(reliability_symbol, tract, rate, rate_moe, cv, reliability) %>%
  gt() %>%
  tab_header(
    title = "Fifteen least reliable tract poverty-rate estimates",
    subtitle = "Philadelphia County, 2013-2017 ACS five-year estimates"
  ) %>%
  cols_label(
    reliability_symbol = "",
    tract = "Census tract",
    rate = "Poverty rate",
    rate_moe = "MOE",
    cv = "CV (%)",
    reliability = "Reliability"
  ) %>%
  fmt_percent(columns = rate, decimals = 1) %>%
  fmt_percent(columns = rate_moe, decimals = 1) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_align(align = "center", columns = reliability_symbol) %>%
  tab_source_note(
    "Source: U.S. Census Bureau ACS five-year estimates via tidycensus. Jurjevich et al. (2018) thresholds: CV < 12% reliable; 12-40% somewhat reliable; > 40% unreliable."
  )
```

### Change over time

Next, I reshape the data so that each tract has one row containing poverty
rates and margins of error for both ACS periods.

```{r}
#| label: reshape-poverty-data

poverty_wide <- poverty_rates_clean %>%
  select(
    GEOID,
    NAME,
    period,
    poverty_rate,
    poverty_rate_moe
  ) %>%
  pivot_wider(
    names_from = period,
    values_from = c(poverty_rate, poverty_rate_moe)
  )
```

I identify tracts that have usable data in only one period. Those tracts cannot
be included in the change test because a difference requires valid estimates
and margins of error for both periods.

```{r}
#| label: inspect-incomplete-comparisons

incomplete_tracts <- poverty_wide %>%
  filter(
    is.na(`poverty_rate_2006-2010`) |
    is.na(`poverty_rate_2013-2017`) |
    is.na(`poverty_rate_moe_2006-2010`) |
    is.na(`poverty_rate_moe_2013-2017`)
  )

incomplete_tracts
```

The ACS reports 90 percent margins of error. I divide each MOE by 1.645 to
calculate its standard error, combine the two standard errors to calculate the
standard error of the difference, and multiply by 1.645 to obtain the
90 percent margin of error for the change.

```{r}
#| label: test-poverty-rate-change

poverty_tested <- poverty_wide %>%
  filter(
    !is.na(`poverty_rate_2006-2010`),
    !is.na(`poverty_rate_2013-2017`),
    !is.na(`poverty_rate_moe_2006-2010`),
    !is.na(`poverty_rate_moe_2013-2017`)
  ) %>%
  mutate(
    change = `poverty_rate_2013-2017` -
      `poverty_rate_2006-2010`,

    se_1 = `poverty_rate_moe_2006-2010` / 1.645,
    se_2 = `poverty_rate_moe_2013-2017` / 1.645,

    se_diff = sqrt(se_1^2 + se_2^2),
    moe_diff = se_diff * 1.645,

    z = change / se_diff,
    significant = abs(z) > 1.645,

    change_direction = case_when(
      significant & change > 0 ~ "Significant increase",
      significant & change < 0 ~ "Significant decrease",
      TRUE ~ "Not distinguishable from zero"
    )
  )

change_counts <- poverty_tested %>%
  count(change_direction)

change_counts
```

Of the `r nrow(poverty_tested)` Census tracts with usable estimates in both
periods, `r sum(poverty_tested$significant)` show a statistically detectable
poverty-rate change at the 90 percent confidence level. The remaining
`r sum(!poverty_tested$significant)` tracts do not show a change that is
statistically distinguishable from zero.

```{r}
#| label: change-summary-plot
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Most Philadelphia census tracts did not have a statistically detectable change in poverty rate."

ggplot(change_counts, aes(x = change_direction, y = n, fill = change_direction)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.15) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Significant increase" = "#C85A5A",
      "Significant decrease" = "#4F81BD",
      "Not distinguishable from zero" = "#8A8A8A"
    )
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Results of the tract-level poverty-rate change test",
    subtitle = "90% confidence tests for 376 matched Philadelphia census tracts",
    x = NULL,
    y = "Number of census tracts",
    caption = "Source: U.S. Census Bureau, ACS five-year estimates via tidycensus."
  ) +
  theme_minimal()
```

```{r}
#| label: change-error-bar-plot
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Largest statistically detectable tract-level poverty-rate changes, with 90% margins of error."

change_plot_data <- poverty_tested %>%
  filter(significant) %>%
  arrange(desc(abs(change))) %>%
  slice_head(n = 20) %>%
  mutate(
    tract = str_remove(NAME, ", Philadelphia County, Pennsylvania"),
    change_pp = change * 100,
    moe_diff_pp = moe_diff * 100,
    direction = case_when(
      change > 0 ~ "Increase",
      TRUE ~ "Decrease"
    )
  )

ggplot(change_plot_data, aes(x = reorder(tract, change_pp), y = change_pp)) +
  geom_col(aes(fill = direction), width = 0.7) +
  geom_errorbar(
    aes(ymin = change_pp - moe_diff_pp, ymax = change_pp + moe_diff_pp),
    width = 0.25
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  scale_fill_manual(values = c("Increase" = "#C85A5A", "Decrease" = "#4F81BD")) +
  labs(
    title = "Largest statistically detectable poverty-rate changes",
    subtitle = "Twenty largest absolute changes, 2006-2010 versus 2013-2017",
    x = NULL,
    y = "Change in poverty rate (percentage points)",
    fill = NULL,
    caption = "Error bars show 90% margins of error; dashed line marks no change. Source: U.S. Census Bureau, ACS five-year estimates via tidycensus."
  ) +
  theme_minimal()
```

## Part 3 — Adjudicate

### Supported

The tract-level analysis supports a limited version of the article's broader
claim that poverty changed in parts of Philadelphia. Among the 376 Census
tracts with usable estimates for both ACS periods, 36 tracts showed a
statistically detectable increase in poverty rate at the 90 percent confidence
level. This result is consistent with the article's general argument that some
places in Philadelphia experienced worsening poverty after the earlier
2006–2010 period.

However, this evidence does not support a claim that poverty increased
everywhere in Philadelphia. Twenty-seven tracts showed statistically detectable
decreases, while 313 tracts had changes that were not distinguishable from zero.
The tract-level evidence therefore indicates a mixed pattern of change rather
than a uniform citywide increase in poverty.

### Not supported

No specific neighborhood-level numeric claim in the article can be classified
as directly not supported by this analysis. A tract with a numerical increase
that is not statistically distinguishable from zero should not be described as
having a confirmed increase. However, because the article reports custom
neighborhood estimates while this analysis uses Census tracts, a tract-level
non-significant result does not directly contradict a neighborhood-level claim.

### Cannot be determined

The article's specific claim that Eastwick had the largest poverty-rate increase
of any Philadelphia neighborhood, rising from 12 percent to 26 percent, cannot
be determined from the available analysis. The article does not publish the
tracts included in Eastwick or the neighborhood boundary used to calculate its
reported rate. As a result, I cannot reproduce the article's exact numerator,
denominator, or margin of error for Eastwick.

The Fairhill claim of a 61 percent poverty rate also cannot be determined for
the same reason: Fairhill is a neighborhood rather than a Census geography.
Similarly, the Holmesburg claim about white poverty cannot be evaluated with
the current analysis because I use overall poverty table `B17001`, not a
race-specific poverty measure. The article also does not define exactly how it
measures “white” residents or publish its neighborhood aggregation method.

## Part 4 — Rewrite

New Census estimates show that poverty changed unevenly across Philadelphia
between 2006–10 and 2013–17. Of 376 Census tracts with comparable estimates,
36 recorded a statistically detectable increase in poverty, while 27 recorded
a detectable decrease. In most tracts, the apparent change could not be
separated from normal survey uncertainty. The data therefore show evidence of
worsening poverty in some places, but not a uniform increase across the city.
They also cannot confirm the article’s exact neighborhood figures, including
Eastwick’s reported rise, because those figures rely on neighborhood boundaries
that were not published.






