---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Chenyan Zhao"
date: September 7, 2026
format:
  html:
    toc: true
    code-fold: false
execute:
  # Keep the unfinished template display-only until its placeholders are filled.
  eval: true
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


In this analysis, I look at **median gross rent** across counties in **New York State**. I compare the margins of error for these estimates to identify which counties’ rent data are measured most and least reliably.

---

## Choosing what to measure

```{r}
my_state    <- "NY"             # two-letter abbreviation, in quotes
my_variable <- "B25064_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

I get 62 rows, which match the number of counties in New York State.

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

Hamilton County has the largest relative margin of error for median gross rent in New York State. Its estimated median gross rent is $860, with a margin of error of $95, or about 11.0% of the estimate.

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

Among New York’s 62 counties, 51 are classified as high confidence, 10 as moderate, and 1 as low confidence.

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```


Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```
I use the 5% and 10% thresholds established in the previous section so that each reliability category has the same meaning throughout the analysis. This ensures that the population comparison uses the same county classifications as the earlier reliability summary.

**High-confidence counties have an average population of about 381,479, while moderate-confidence counties average about 41,178. The only low-confidence county, Hamilton County, has a population of 5,102. This pattern suggests that median gross rent estimates are less reliable in counties with smaller populations, although the low-confidence category contains only one county.**

---

## What this means for policy

**Which communities in your state do we know the least about?**

This analysis shows that New York’s smallest counties tend to have the least reliable median gross rent estimates.   Hamilton County is the clearest example: its estimated median gross rent is $860 with a margin of error of $95, or about 11.0 percent. It is the only low-confidence county under the 10 percent cutoff and has only 5,102 residents. Schuyler County has the next-highest relative margin of error at 8.7 percent, followed by Greene County at 8.0 percent.



**Why?**

These differences reflect how the American Community Survey works. The ACS draws estimates from a sample rather than counting every household. Smaller counties generally contain fewer sampled households, so a small number of responses can have a larger effect on their estimates. In these results, high-confidence counties average about 381,479 residents, compared with 41,178 in moderate-confidence counties and 5,102 in the low-confidence county. This pattern indicates that estimates become less precise as county population decreases.

**What could go wrong?**

If a state agency allocated housing assistance using median gross rent while treating every estimate as equally certain, small rural counties could be harmed. An underestimated rent could make a county appear less burdened and cause it to receive too little funding, while an overestimate could divert funds from places with greater need. Renters in Hamilton, Schuyler, and Greene Counties may therefore be especially vulnerable. Agencies should consider both the estimates and their margins of error, and they should avoid making major funding decisions based on small differences that may reflect sampling uncertainty.
